Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-08 Manuscript/manuscript.Rmd
edits to methods (enriched heatmap details) and results (paragraph 1)

We first established a *predicted TAD boundary regions* (PTBR) as clusters of densitity-reachable base pairs which have been identified as being a predicted TAD boundary with probability 1 according to the random forest model using *DBSCAN* (version 1.1-5) [@hahsler2019dbscan]. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The eps parameter ($\epsilon$) was assigned as the respective data resolution. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. [??? This deserves a schematic figure, before Figure 5. And part of it may be in Results;!!!added PTBR to example of Figure 5 A] [??? I would switch the notation. First, you detect REGIONS. They can be adjacent, clustered together, forming AREA. Within this area, you can do pam clustering, and identify POINT;!!!editted, used term "region" instead of "area"]

To precisely identify the PTBA most representative base pair of each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The midpoint of the corresponding cluster medoid, in this case a particular PTBR, was defined as a *preciseTAD boundary point* (PTBP). <!--Thus, the hierarchical structure of our defined predicted TAD boundary objects were as follows: PTBR > PTBA > PTBP.--> That is, each PTBR is a collection of base pairs, with the uniquely identified PTBP being the most representative base pair.

@@

Next, we evaluated the biological significance of our predicted boundaries with respect to the enrichment of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. Enrichmed heatmaps were used to compare preciseTAD boundaries to both ARROWHEAD and PEAKACHU called boundaries using the `EnrichedHeatmap` R package (version 1.14.0) [@gu2018enrichedheatmap]. Furthermore, we compared the $log_2$ [??? log2?] distance between TAD boundaries and the same top predictive transcription factor binding sites using Wilcoxon Rank-Sum test.

Under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell types and prediction algorithms. Boundaries were flanked by resolution and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) [??? Package?;!!!added] and evaluated using the Jaccard index defined as

@@

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. Random forest machine learning algorithm was selected as the most robust to overfitting [??? Read 10.1002/widm.1072 and 10.1002/widm.1301, elaborate, cite] We then leveraged our optimally built random forest to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called [??? Did you call them?] using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]

>>>>

We first established *predicted TAD boundary regions* (PTBR) as clusters of density-reachable base pairs which have been identified as being a predicted TAD boundary with probability 1 according to the random forest model using *DBSCAN* (version 1.1-5) [@hahsler2019dbscan]. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The eps parameter ($\epsilon$) was assigned as the respective data resolution. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3. [??? This deserves a schematic figure, before Figure 5. And part of it may be in Results;!!!added PTBR to example of Figure 5 A] [??? I would switch the notation. First, you detect REGIONS. They can be adjacent, clustered together, forming AREA. Within this area, you can do pam clustering, and identify POINT;!!!editted, used term "region" instead of "area"] To precisely identify the most representative base pair among each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The midpoint of the corresponding cluster medoid, in this case a particular PTBR, was defined as a *preciseTAD boundary point* (PTBP). <!--Thus, the hierarchical structure of our defined predicted TAD boundary objects were as follows: PTBR > PTBA > PTBP.--> That is, each PTBR is a cluster of base pairs highly probable of being a boundary, with the uniquely identified PTBP being the most representative base pair.

@@

Next, we evaluated the biological significance of our predicted boundaries with respect to the enrichment of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. Enrichmed heatmaps were used to compare preciseTAD boundaries to both ARROWHEAD and PEAKACHU called boundaries using the `EnrichedHeatmap` R package (version 1.14.0) [@gu2018enrichedheatmap]. For each boundary (both called and predicted), a 5000 bp flank was created on either side. The flanking was broken up into 50 bp segments, for a total of 100 windows on both sides of a given boundary. A coverage value was generated for each window with respect to a given ChIP-seq annotation, as the mean signal averaged by the width of the window, denoted by,

$$v_{c} = \dfrac{\Sigma_{j}^{m} \Sigma_{i}^{n}x_{ij}w_{ij}}{L}$$
where $x_{ij}$ is the total number of base pairs of annotation $i$ in window $j$ and $w_{ij}$ is the number of base pairs that overlap between annotation $i$ and window $j$. The denominator, $L$, is the width of the windows (here, $L=50$). The rows of the heatmaps are ordered by enriched scores. The enriched scores are calculated as the sum of coverage values weighted by the distance to the flanked boundary, denoted by

$$s_{e} = \Sigma_{d=1}^{n_{1}}\left(\dfrac{x_{1d}* d}{n_{1}} \right) + \Sigma_{u=1}^{n_{2}}\left(\dfrac{x_{2u}* (n_{2} - u + 1)}{n_{2}} \right)$$
where $n_{1}$ and $n_{2}$ represent the number of downstream and upstream coverage values to sum over. Here, $n_{1}=n_{2}=100$.

Additionally, we compared the $log_2$ [??? log2?;!!!added] distance between TAD boundaries and the same top predictive transcription factor binding sites using Wilcoxon Rank-Sum test. The alpha level of significance was set to 0.05.

Furthermore, under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell types and prediction algorithms. Boundaries were flanked by resolution and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) [??? Package?;!!!added] and evaluated using the Jaccard index defined as

@@

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction. We chose the random forest (RF) machine learning algorithm as our binary classification tool. The reason for choosing RF was two-fold: (1) In order to devise a tunable prediction rule in a supervised learning framework that was both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors with respect to their ability to predict boundary regions [@boulesteix2012overview] [??? Read 10.1002/widm.1072 and 10.1002/widm.1301, elaborate, cite;!!!added first source]. We then leveraged our optimally built RF to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called [??? Did you call them?; !!!yes] using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results]
